\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=1in}

% Code listing style
\lstset{
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue},
    commentstyle=\color{green!50!black},
    stringstyle=\color{red},
    numbers=left,
    numberstyle=\tiny,
    frame=single,
    breaklines=true,
    captionpos=b
}

\title{4DT906 Assignment 1: Parallel Computing\\
Odd-Even Sort and Pearson Correlation Coefficient (C/C++)}
\author{Samuel Berg, Jesper Wingren, Emil Ulvagården}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This report presents parallel implementations of two algorithms: the odd-even transposition sort
and the Pearson correlation coefficient (PCC). Each algorithm is implemented in two ways---using
C++ \texttt{std::thread} for shared-memory parallelism and using MPI for distributed-memory
parallelism. Correctness is validated against the provided sequential reference implementations.
Performance results and speedup measurements are to be filled in after benchmarking.
\end{abstract}

\tableofcontents
\newpage

\section{Introduction}

Parallel computing enables large-scale computations to complete in less time by distributing work
across multiple execution units. This assignment explores two algorithms that are amenable to
parallelisation:

\begin{enumerate}
    \item \textbf{Odd-Even Transposition Sort}: A comparison-based sorting algorithm with a
          regular, phase-alternating communication pattern that maps naturally to both shared- and
          distributed-memory models.
    \item \textbf{Pearson Correlation Coefficient (PCC)}: A statistical measure of linear
          correlation between pairs of data rows. Computing all $\frac{m(m-1)}{2}$ pair-wise
          correlations for an $m \times n$ matrix is computationally intensive and highly
          parallelisable.
\end{enumerate}

For each algorithm a correct sequential reference implementation is provided
(\texttt{oddevensort.cpp} and \texttt{pcc\_seq.cpp}). Parallel versions were developed on top of
these references without modifying them.

\section{Benchmarking Setup}

\begin{itemize}
    \item \textbf{Hardware}: \textbf{[TODO: fill in CPU model, core count, RAM]}
    \item \textbf{Operating System}: \textbf{[TODO: fill in OS and kernel version]}
    \item \textbf{Compiler}: g++ with \texttt{-O2} optimisation, C++17 standard; gcc for the MPI C file
    \item \textbf{MPI implementation}: \textbf{[TODO: e.g.\ OpenMPI 4.x]}
    \item \textbf{Thread counts tested}: \textbf{[TODO: e.g.\ 1, 2, 4, 8, \ldots]}
    \item \textbf{Number of runs per configuration}: \textbf{[TODO: e.g.\ 3 runs, median reported]}
\end{itemize}

\section{Odd-Even Transposition Sort}

\subsection{Algorithm Description}

The odd-even transposition sort alternates between two phases until the array is sorted:

\begin{itemize}
    \item \textbf{Even phase}: compare-and-swap all pairs $(v[0],v[1]),\,(v[2],v[3]),\,\ldots$
    \item \textbf{Odd phase}: compare-and-swap all pairs $(v[1],v[2]),\,(v[3],v[4]),\,\ldots$
\end{itemize}

After $n$ full iterations (even + odd) the array is guaranteed to be sorted.
Within each phase all active pairs are independent, making the algorithm naturally parallel.

\subsection{Sequential Implementation}

The reference implementation (\texttt{oddevensort.cpp}) performs $n$ outer iterations and
processes each active pair in the inner loop:

\begin{lstlisting}[language=C++, caption=Sequential Odd-Even Sort (core loop)]
for (int i = 1; i <= s; i++) {
    for (int j = i % 2; j < s-1; j = j + 2) {
        if (numbers[j] > numbers[j + 1]) {
            std::swap(numbers[j], numbers[j + 1]);
        }
    }
}
\end{lstlisting}

\textbf{Time complexity}: $O(n^2)$ in the worst case.

\subsection{Parallel Implementation --- \texttt{std::thread} (\texttt{oddevensort\_par.cpp})}

\subsubsection{Parallelisation Strategy}

Within each phase the $\lfloor n/2 \rfloor$ active compare-swap pairs are independent. The pairs
are divided into contiguous blocks, one block per thread, so threads operate on disjoint memory
locations and no data races occur. A reusable spin-wait barrier synchronises all threads between
phases, ensuring phase $k+1$ does not begin before phase $k$ has finished.

\subsubsection{Synchronisation}

A custom \texttt{Barrier} class based on \texttt{std::atomic<int>} provides the inter-phase
synchronisation. Each thread calls \texttt{arrive\_and\_wait()} at the end of its phase work; the
last arriving thread resets the counter and increments a generation counter, waking the others.

\subsubsection{Data Decomposition}

Pairs are divided into contiguous blocks to maximise cache locality. Thread $t$ handles pairs
\texttt{[pair\_start, pair\_end)} where the block boundaries are computed to distribute the
remainder evenly:

\begin{lstlisting}[language=C++, caption=Contiguous pair-block assignment]
int per        = num_pairs / num_threads;
int rem        = num_pairs % num_threads;
int pair_start = thread_id * per + (thread_id < rem ? thread_id : rem);
int pair_end   = pair_start + per + (thread_id < rem ? 1 : 0);
\end{lstlisting}

\subsection{Parallel Implementation --- MPI (\texttt{oddevensort\_par\_mpi.cpp})}

\subsubsection{Parallelisation Strategy}

The array is distributed evenly across $p$ MPI processes using \texttt{MPI\_Scatter}. Each
process sorts its local chunk independently, then $p$ rounds of odd-even compare-exchange are
performed between neighbouring processes. Finally the sorted chunks are collected with
\texttt{MPI\_Gather}.

\subsubsection{Compare-Exchange Operations}

Two helper functions implement the inter-process compare-exchange:
\begin{itemize}
    \item \texttt{compare\_exchange\_low}: process retains the smaller half of the merged pair
          of chunks.
    \item \texttt{compare\_exchange\_high}: process retains the larger half.
\end{itemize}

Both functions use \texttt{MPI\_Sendrecv} for simultaneous bidirectional communication:

\begin{lstlisting}[language=C++, caption=MPI compare-exchange (low variant)]
MPI_Sendrecv(local_numbers.data(), n_local, MPI_INT, partner, 0,
             recv_numbers.data(), n_local, MPI_INT, partner, 0,
             MPI_COMM_WORLD, MPI_STATUS_IGNORE);
// merge and keep lower half
\end{lstlisting}

\subsection{Correctness Validation}

Both parallel implementations are verified with \texttt{std::is\_sorted} after execution.
The default input size is $2^{19} = 524\,288$ elements; additional sizes were also tested.

\subsection{Performance Results --- Odd-Even Sort}

\begin{table}[h]
\centering
\begin{tabular}{|r|r|r|r|r|}
\hline
\textbf{Input size} & \textbf{Threads / Procs} & \textbf{Sequential (s)} & \textbf{Parallel (s)} & \textbf{Speedup} \\
\hline
$2^{19}$ & - & - & - & - \\
$2^{19}$ & - & - & - & - \\
$2^{19}$ & - & - & - & - \\
\hline
\end{tabular}
\caption{Odd-Even Sort --- \texttt{std::thread} implementation}
\end{table}

\begin{table}[h]
\centering
\begin{tabular}{|r|r|r|r|r|}
\hline
\textbf{Input size} & \textbf{Threads / Procs} & \textbf{Sequential (s)} & \textbf{Parallel (s)} & \textbf{Speedup} \\
\hline
$2^{19}$ & - & - & - & - \\
$2^{19}$ & - & - & - & - \\
$2^{19}$ & - & - & - & - \\
\hline
\end{tabular}
\caption{Odd-Even Sort --- MPI implementation}
\end{table}

\section{Pearson Correlation Coefficient}

\subsection{Algorithm Description}

The Pearson correlation coefficient between two vectors $x$ and $y$ of length $n$ is:
$$r = \frac{\sum_{k=1}^{n}(x_k - \bar{x})(y_k - \bar{y})}{\sqrt{\sum_{k}(x_k-\bar{x})^2}\,\sqrt{\sum_{k}(y_k-\bar{y})^2}}$$

For a matrix with $m$ rows and $n$ columns, all $\frac{m(m-1)}{2}$ row-pair correlations are
computed. The calculation proceeds in three steps:
\begin{enumerate}
    \item Compute the row mean $\bar{x}_i$ for every row $i$.
    \item Compute the mean-subtracted matrix $(x_{ik} - \bar{x}_i)$ and the row standard
          deviation $\sigma_i$.
    \item Compute each pair-wise correlation
          $r_{ij} = \frac{\sum_k (x_{ik}-\bar{x}_i)(x_{jk}-\bar{x}_j)}{\sigma_i\,\sigma_j}$.
\end{enumerate}

\subsection{Sequential Implementation}

The reference implementation (\texttt{pcc\_seq.cpp}) computes all three steps in sequence. The
correlation loop is:

\begin{lstlisting}[language=C++, caption=Sequential PCC correlation loop]
for (sample1 = 0; sample1 < ROWS-1; sample1++) {
    for (sample2 = sample1+1; sample2 < ROWS; sample2++) {
        sum = 0.0;
        for (i = 0; i < COLS; i++) {
            sum += mm[sample1*COLS+i] * mm[sample2*COLS+i];
        }
        r = sum / (std[sample1] * std[sample2]);
        output[...] = r;
    }
}
\end{lstlisting}

\textbf{Time complexity}: $O(m^2 n)$.

\subsection{Parallel Implementation --- \texttt{std::thread} (\texttt{pcc\_par.cpp})}

\subsubsection{Parallelisation Strategy}

All three steps (mean, deviation/std, correlations) are parallelised. Steps 1 and 2 decompose
the work by rows: each thread receives a contiguous slice of rows. Step 3 uses interleaved
(round-robin) assignment of \texttt{sample1} rows across threads, which distributes the
triangular workload evenly because rows with a higher index produce more pairs.

\subsubsection{Data Decomposition}

\begin{itemize}
    \item \textbf{Steps 1 \& 2 (linear)}: rows $[0, m)$ are split into contiguous blocks of
          size $\lceil m/T \rceil$ using a helper lambda.
    \item \textbf{Step 3 (triangular)}: thread $t$ processes \texttt{sample1} rows
          $t,\, t+T,\, t+2T,\,\ldots$ (stride $T$). Because each output index
          \texttt{sample1*ROWS + sample2 - offset} is written by exactly one thread, no
          synchronisation is required inside the correlation loop.
\end{itemize}

\subsubsection{Synchronisation}

No explicit synchronisation is needed within steps 1--3 because the work partitions are
non-overlapping. Thread joining (\texttt{th.join()}) serves as the implicit barrier between
steps.

\subsection{Parallel Implementation --- MPI (\texttt{pcc\_par\_mpi.c})}

\subsubsection{Parallelisation Strategy}

All processes generate the same input matrix (avoiding a broadcast). Steps 1 and 2 (mean and
standard deviation) are computed redundantly by every process; the cost is $O(mn)$ and small
compared to step 3. Step 3 distributes the $\frac{m(m-1)}{2}$ correlation pairs evenly among
processes: process $r$ computes pairs in the global index range
$[\textit{local\_start},\,\textit{local\_end})$.

\subsubsection{Work Distribution}

\begin{lstlisting}[language=C, caption=MPI work distribution for PCC]
int local_size  = cor_size / size;
int remainder   = cor_size % size;
int local_start = rank * local_size +
                  (rank < remainder ? rank : remainder);
int local_count = local_size + (rank < remainder ? 1 : 0);
int local_end   = local_start + local_count;
\end{lstlisting}

Results are collected at the root with \texttt{MPI\_Gatherv}, which handles the unequal chunk
sizes arising from the remainder.

\subsection{Correctness Validation}

The provided \texttt{verify} program compares the floating-point output files of the sequential
and parallel implementations. Both parallel versions are validated for all tested matrix sizes.

\subsection{Performance Results --- Pearson Correlation Coefficient}

\begin{table}[h]
\centering
\begin{tabular}{|r|r|r|r|r|}
\hline
\textbf{Matrix size} & \textbf{Threads} & \textbf{Sequential (s)} & \textbf{Parallel (s)} & \textbf{Speedup} \\
\hline
$64 \times 64$     & - & - & - & - \\
$128 \times 128$   & - & - & - & - \\
$256 \times 256$   & - & - & - & - \\
$512 \times 512$   & - & - & - & - \\
$1024 \times 1024$ & - & - & - & - \\
$2048 \times 2048$ & - & - & - & - \\
$4096 \times 4096$ & - & - & - & - \\
\hline
\end{tabular}
\caption{PCC --- \texttt{std::thread} implementation}
\end{table}

\begin{table}[h]
\centering
\begin{tabular}{|r|r|r|r|r|}
\hline
\textbf{Matrix size} & \textbf{Processes} & \textbf{Sequential (s)} & \textbf{Parallel (s)} & \textbf{Speedup} \\
\hline
$64 \times 64$     & - & - & - & - \\
$128 \times 128$   & - & - & - & - \\
$256 \times 256$   & - & - & - & - \\
$512 \times 512$   & - & - & - & - \\
$1024 \times 1024$ & - & - & - & - \\
$2048 \times 2048$ & - & - & - & - \\
$4096 \times 4096$ & - & - & - & - \\
\hline
\end{tabular}
\caption{PCC --- MPI implementation}
\end{table}

\section{Discussion}

\subsection{Scalability and Observed Behaviour}

\textbf{[TODO: fill in after benchmarking --- describe speedup trends, crossover points, and any surprising results]}
\begin{itemize}
    \item \textbf{Odd-Even Sort}: The algorithm performs $n$ phases with a barrier between each
          phase. For the thread-based version, synchronisation overhead grows with thread count,
          potentially limiting scalability for small arrays. For the MPI version, inter-process
          communication cost grows with process count.
    \item \textbf{PCC}: The correlation step dominates runtime for large matrices, making PCC
          well-suited for parallelisation. Both implementations are expected to scale well for
          large matrices where computation time dominates overhead.
    \item \textbf{Communication vs.\ Computation}: For small inputs, thread/process management
          overhead and communication costs can exceed the computational savings, leading to
          reduced or negative speedup. As problem size increases, the compute-to-overhead ratio
          improves.
\end{itemize}

\subsection{Memory Behaviour}

\begin{itemize}
    \item \textbf{Odd-Even Sort (threads)}: Contiguous pair-block assignment improves spatial
          locality compared to interleaved assignment. Each thread accesses a contiguous region,
          reducing cache-line contention.
    \item \textbf{Odd-Even Sort (MPI)}: The compare-exchange operation requires a temporary
          merge buffer of size $2 \cdot n_{\text{local}}$ per process during each exchange,
          doubling peak memory usage.
    \item \textbf{PCC (threads)}: Each thread reads the full mean-subtracted matrix rows it
          needs; rows are accessed sequentially, maintaining good cache behaviour.
    \item \textbf{PCC (MPI)}: All processes allocate the full $m \times n$ input matrix,
          which may limit scalability on memory-constrained nodes for very large matrices.
\end{itemize}

\subsection{Load Balancing}

\begin{itemize}
    \item \textbf{Odd-Even Sort}: The per-thread workload is perfectly balanced since each
          active phase has the same number of pairs and these are distributed evenly.
    \item \textbf{PCC (threads)}: The interleaved (round-robin) row assignment in step 3
          distributes the triangular workload evenly across threads, avoiding the imbalance that
          would arise from a simple contiguous partition.
    \item \textbf{PCC (MPI)}: Correlation pairs are divided by global index; the remainder is
          distributed one pair at a time to the first \texttt{remainder} processes, giving at
          most one extra pair per process.
\end{itemize}

\section{Conclusion}

Both odd-even sort and the Pearson correlation coefficient were parallelised successfully using
two complementary models: C++ \texttt{std::thread} for shared-memory execution and MPI for
distributed-memory execution. Correctness was validated against the provided sequential
reference programs.

Performance results and speedup values will be added after benchmarking on the target hardware.
The implementations are expected to achieve positive speedup ($>2\times$) for non-trivial input
sizes, with efficiency improving as problem size grows and the compute-to-overhead ratio
increases.

\section*{References}

\begin{itemize}
    \item MPI Forum. \textit{MPI: A Message-Passing Interface Standard, Version 3.1}. 2015.
    \item Quinn, M.~J. \textit{Parallel Programming in C with MPI and OpenMP}. McGraw-Hill, 2004.
    \item Course material: \texttt{pa1\_26.pdf} --- Programming Assignment 1 instructions.
\end{itemize}

\end{document}
